---
output:
  md_document:
    variant: markdown_github
---
[![Build Status](https://travis-ci.org/jamesdunham/dgirt.svg?branch=master)](https://travis-ci.org/jamesdunham/dgirt)

dgirt is an R package for dynamic group-level item response theory (DGIRT)
models.

DGIRT is a Bayesian method for estimating subpopulation groups' average
conservatism (or other trait) from individuals' responses to dichotomous
questions. It is "dynamic" both in the sense that groups are allowed to evolve
over time and in the sense that the model "borrows strength" from other time
periods, to a degree specified by the user. [This
document](https://github.com/jamesdunham/dgirt/blob/master/man/dgirt_details.pdf)
describes the model in detail.

It is a modified version of the hierarchical group-level IRT model
implemented by [Caughey and Warshaw 2015](http://pan.oxfordjournals.org/content/early/2015/02/04/pan.mpu021.full.pdf+html).

```{r, knitr-options, echo = FALSE}
# rmarkdown::render("README.Rmd")
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  cache = FALSE)
```

## Prerequisites

Installation requires [RStan](http://mc-stan.org/interfaces/rstan.html) and its
prerequisites, in particular a C++ toolchain. If you don't have RStan, follow
its "[Getting
Started](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started)" guide
before continuing.

## Installation

dgirt can be installed from [GitHub](https://github.com/jamesdunham/dgirt) using
[devtools](https://github.com/hadley/devtools/):
```
devtools::install_github("jamesdunham/dgirt", dependencies = TRUE)
```

## Getting started

```{r}
library(dgirt)
```

The minimal workflow from raw data to estimation is:

  1. shape input data using the `shape` function; and
  2. pass the result to the `dgirt` function to fit a dgirt model.

### Set RStan options

These are RStan's recommended options on a local, multicore machine with excess
RAM:

```{r}
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

### Prepare input data with `shape`

dgirt models are *dynamic*, so we need to specify which variable in the data
represents time. They are also *group-level*, with groups defined by one
variable for respondents' local geographic area and one or more variables for
respondent characteristics.

The `time_filter` and `geo_filter` arguments optionally subset the data.
Finally, `shape` requires the names of the survey identifier and survey weight
variables in the data.

```{r}
dgirt_in <- shape(opinion,
                  item_names = "Q_cces2006_abortion",
                  time_name = "year",
                  geo_name = "state",
                  group_names = "race",
                  geo_filter = c("CA", "GA", "LA", "MA"),
                  survey_name = "source",
                  weight_name = "weight")
```

The reshaped and subsetted data can be summarized in a few ways before model
fitting.

```{r}
summary(dgirt_in)
```

Response counts by survey-year:
```{r}
get_n(dgirt_in, by = c("year", "source"))
```

Response counts by item-year:
```{r}
get_item_n(dgirt_in, by = "year")
```

### Fit a model with `dgirt`

`dgirt` fits a model to data from `shape`. Under the hood, this function uses
RStan for MCMC sampling, and arguments can be passed to RStan's `stan` via the
`...` argument of `dgirt`. This will almost always be desirable, at a minimum to
specify the number of sampler iterations, chains, and cores.

```{r, warning = FALSE, cache = TRUE}
dgirt_out <- dgirt(dgirt_in, iter = 1500, chains = 4, cores = 4, seed = 42,
                   refresh = 0)
```

The model results are held in a `dgirtfit` object. Methods from RStan like
`extract` are available if needed because `dgirtfit` is a subclass of `stanfit`.
But dgirt provides its own methods for typical post-estimation tasks.

### Work with `dgirt` results

For a high-level summary of the result, use `summary`.

```{r}
summary(dgirt_out)
```

To summarize posterior samples, use `summarize`. The default output gives
summary statistics for the `theta_bar` parameters, which represent the mean of
the latent outcome for the groups defined by time, local geographic area, and
the demographic characteristics specified in the earlier call to `shape`.

```{r}
head(summarize(dgirt_out))
```

Alternatively, `summarize` can apply arbitrary functions to posterior samples
for whatever parameter is given by its `pars` argument. Enclose function names
with quotes. For convenience, `"q_025"` and `"q_975"` give the 2.5th and 97.5th
posterior quantiles.

```{r}
summarize(dgirt_out, pars = "xi", funs = "var")
```

To access posterior samples in tabular form use `as.data.frame`. By default,
this method returns post-warmup samples for the `theta_bar` parameters, but like
other methods takes a `pars` argument.

```{r}
head(as.data.frame(dgirt_out))
```

To poststratify the results use `poststratify`. The following example uses the
group population proportions bundled as `state_year_targets` to reweight and
aggregate estimates to strata defined by state-years. Read
`help("poststratify")` for more details.

```{r}
poststratify(dgirt_out, state_year_targets, strata_names = c("state", "year"),
             aggregated_names = "race")
```

To plot the results use `dgirt_plot`. This method plots summaries of posterior
samples by time period. By default, it shows a 95% credible interval around
posterior medians for the `theta_bar` parameters, for each local geographic
area. For this (unconverged) toy example we omit the CIs.

```{r}
dgirt_plot(dgirt_out, y_min = NULL, y_max = NULL)
```

Output from `dgirt_plot` can be customized to some extent using objects from the
ggplot2 package.

```{r}
dgirt_plot(dgirt_out, y_min = NULL, y_max = NULL) + theme_classic()
```

`dgirt_plot` can also plot the `data.frame` output from `poststratify`. This
requires arguments that identify the relevant variables in the `data.frame`.
Below, `poststratify` aggregates over the demographic grouping variable `race`,
resulting in a `data.frame` of estimates by state-year. So, in the subsequent
call to `dgirt_plot`, we pass the names of the state and year variables. The
`group_names` argument is `NULL` because there are no grouping variables left
after aggregating over `race`.

```{r}
ps <- poststratify(dgirt_out, state_year_targets,
                   strata_names = c("state", "year"), aggregated_names = "race")
head(ps)
dgirt_plot(ps, group_names = NULL, time_name = "year", geo_name = "state")
```

## Troubleshooting

Please [report issues](https://github.com/jamesdunham/dgirt/issues) that you
encounter.

OS X only: RStan creates temporary files during estimation in a location given
by `tempdir`, typically an arbitrary location in `/var/folders`. If a model runs
for days, these files can be cleaned up while still needed, which induces an
error. A good solution is to set a safer path for temporary files, using an
environment variable checked at session startup. As described in `?tempdir`, 

> The environment variables ‘TMPDIR’, ‘TMP’ and ‘TEMP’ are checked in turn and
> the first found which points to a writable directory is used: if none succeeds
> ‘/tmp’ is used. The path should not contain spaces.  

For help setting environment variables, see the Stack Overflow question
[here](https://stackoverflow.com/questions/17107206/change-temporary-directory).
Confirm the new path before starting your model run by restarting R and checking
the output from `tempdir()`.

``` r
# Problematic temporary directories on OS X look like this
tempdir()   
#> [1] "/var/folders/2p/_d3c95qd6ljg28j1f5l2jqxm0000gn/T//Rtmp38a10A"
```

## Contributing and citing

dgirt is under development and we welcome
[suggestions](https://github.com/jamesdunham/dgirt/issues).

The package citation is 

> Dunham, James, Devin Caughey, and Christopher Warshaw. 2016. dgirt: Dynamic
> Group-level IRT Models in R. R package. https://jamesdunham.github.io/dgirt/.
