---
output:
  md_document:
    variant: markdown_github
---
[![Build Status](https://travis-ci.org/jamesdunham/dgirt.svg?branch=master)](https://travis-ci.org/jamesdunham/dgirt)

dgirt is an R package for dynamic group-level IRT models as developed in
[Caughey and Warshaw 2014](http://pan.oxfordjournals.org/content/early/2015/02/04/pan.mpu021.full.pdf+html).

```{r, knitr-options, echo = FALSE}
# rmarkdown::render("README.Rmd")
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  cache = FALSE)
```

## Install

Install dgirt from GitHub. Installation requires
[RStan](http://mc-stan.org/interfaces/rstan.html) and its prerequisites, in
particular a C++ toolchain. If you don't have RStan, follow its "[Getting
Started](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started)" guide.

```
devtools::install_github("jamesdunham/dgirt", dependencies = TRUE)
```

RStan's recommended options on a local, multicore machine with excess RAM are
also appropriate for dgirt:

```{r}
library(dgirt)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

## Getting started

The minimal workflow from raw data to estimation is:

  1. shape input data using the `shape` function; and
  2. pass the result to the `dgirt` function to fit a dgirt model.

### Prepare input data with `shape`

The following example models responses to a single survey item in the bundled
`opinion` data. dgirt models are *dynamic*, so we need to specify which variable
in the data represents time. They are also *group-level*, with groups defined by
one variable for respondents' local geographic area and one or more variables
for respondent characteristics.

The `time_filter` and `geo_filter` arguments optionally subset the example data.
`time_filter` may include unobserved periods. Finally, `shape` requires the
names of the survey identifier and survey weight variables in the data.

```{r}
dgirt_in <- shape(opinion, item_names = "Q_cces2006_abortion",
                  time_name = "year", geo_name = "state", group_names = "race",
                  time_filter = 2006:2008, geo_filter = c("MA", "NY"),
                  survey_name = "source", weight_name = "weight")
```

The reshaped and subsetted data can be summarized in a few ways before model
fitting.

```{r}
summary(dgirt_in)
get_n(dgirt_in, by = c("year", "source"))
get_item_n(dgirt_in, by = "year")
```

### Fit a model with `dgirt`

We use `dgirt` to fit a model to data from `shape`. Under the hood, this
function uses RStan for MCMC sampling, and arguments can be passed to RStan's
`stan` via the `...` argument of `dgirt`. This will almost always be desirable,
at a minimum to specify the number of sampler iterations, chains, and cores.

```{r, warning = FALSE}
dgirt_out <- dgirt(dgirt_in, iter = 2000, chains = 4, cores = 4, seed = 42,
                   refresh = 0)
```

The model results are held in a `dgirtfit` object. This means that methods from
RStan like `extract` are available if needed. But dgirt provides its own methods
for typical post-estimation tasks.

### Work with `dgirt` results

For a high-level summary of the result, use `summary`.

```{r}
summary(dgirt_out)
```

To summarize posterior samples, use `summarize`. The default output gives
summary statistics for the `theta_bar` parameters, which represent the mean of
the latent outcome for the groups defined by time, local geographic area, and
the demographic characteristics specified in the earlier call to `shape`.

```{r}
summarize(dgirt_out)
```

Alternatively, `summarize` can apply arbitrary functions to posterior samples
for whatever parameter is given by its `pars` argument. Enclose function names
with quotes. For convenience, `"q_025"` and `"q_975"` give the 2.5th and 97.5th
posterior quantiles.

```{r}
summarize(dgirt_out, pars = "xi", funs = "var")
```

To access posterior samples in tabular form use `as.data.frame`. By default,
this method returns post-warmup samples for the `theta_bar` parameters, but like
other methods takes a `pars` argument.

```{r}
as.data.frame(dgirt_out)
```

To poststratify the results use `poststratify`. This example uses the group
population proportions bundled as `state_year_targets` to reweight and aggregate
estimates to strata defined by state-years. Read `help("poststratify")` for more
details. 

```{r}
poststratify(dgirt_out, state_year_targets, strata_names = c("state", "year"),
             aggregated_names = "race")
```

To plot the results use `dgirt_plot`. This method plots summaries of posterior
samples by time period. By default, it shows a 95% credible interval around
posterior medians for the `theta_bar` parameters, for each local geographic
area. 

```{r}
dgirt_plot(dgirt_out)
```

Output from `dgirt_plot` can be customized using objects from the ggplot2
package.

```{r}
dgirt_plot(dgirt_out) + theme_classic()
```
