---
output: github_document
---

# Abortion Attitudes

This vignette demonstrates estimation of public attitudes toward abortion from
responses to a single survey item, using the dynamic multi-level regression and
post-stratification (MRP) model implemented in `dgmrp()`.

```{r}
library(dgo)
```

## Prepare input data with `shape()`

`shape()` prepares input data for use with the modeling functions `dgirt()` and
`dgmrp()`. Here we use the included `opinion` dataset.

```{r, message = FALSE, result = 'hide'}
dgirt_in_abortion <- shape(opinion,
  item_names = "abortion",
  time_name = "year",
  geo_name = "state",
  group_names = "race3",
  geo_filter = c("CA", "GA", "LA", "MA"),
  id_vars = "source")
```

In this call to `shape()` we specified:

* the survey item response variable (`abortion`);
* which variable represents time (`year`), since dgo models are dynamic;
* the variables representing respondent characteristics (`state` and `race3`),
  because dgo models are group-level.

Notice that we named only one of these variables defining respondent groups
using the `group_names` argument. The `geo_name` argument always takes the
variable giving respondents' local geographic area; it will be modeled
differently.

Using the argument `geo_filter`, we subset the input data to the given values of
the `geo_name` variable. And with the `id_vars` argument, we named an identfier
that we'd like to keep in the processed data. (Other unused variables will be
dropped.)

## Inspect the result

`summary()` gives a high-level description of the result.

```{r}
summary(dgirt_in_abortion)
```

`get_n()` and `get_item_n()` give response counts.

```{r}
get_n(dgirt_in_abortion, by = "state")
get_item_n(dgirt_in_abortion, by = "year")
```

## Fit a model with `dgmrp()`

`dgmrp()` fits a dynamic multi-level regression and post-stratification (MRP) model to data processed by `shape()`. Here, we'll use it to estimate public attitudes toward abortion over time, for the groups defined by `state` and `race3`. (Specifically, by their Cartesian product.)

Under the hood, `dgmrp()` uses RStan for MCMC sampling, and arguments can be
passed to RStan's
[`stan()`](https://www.rdocumentation.org/packages/rstan/versions/2.15.1/topics/stan)
via the `...` argument of `dgmrp()`. This is almost always desirable. Here, we
specify the number of sampler iterations, chains, and cores.

```{r, warning = FALSE, message = FALSE, results = 'hide'}
dgmrp_out_abortion <- dgmrp(dgirt_in_abortion, iter = 1500, chains = 4, cores =
  4, seed = 42)
```

## Work with `dgmrp()` results

The results are held in a `dgmrp_fit` object. This class inherits from RStan's
`stanfit`, so post-estimation methods from that package are available. Below, we
demonstrate dgo's custom methods.

For a high-level summary of the result, use `summary`.

```{r}
summary(dgmrp_out_abortion)
```

To apply scalar functions to posterior samples, use
[`summarize()`](https://jdunham.io/dgo/reference/dgo_fit-methods.html). The
default output gives summary statistics for the model's `theta_bar` parameters,
which represent group means. These are indexed by time (`year`) and group, where
groups are again defined by local geographic area (`state`) and any other
respondent characteristics (`race3`).

```{r}
head(summarize(dgmrp_out_abortion))
```

Alternatively,
[`summarize()`](https://jdunham.io/dgo/reference/dgo_fit-methods.html) can apply
arbitrary functions to posterior samples for whatever parameter is given by its
`pars` argument.

```{r}
summarize(dgmrp_out_abortion, pars = "xi", funs = "var")
```

To access posterior samples in tabular form use [`as.data.frame()`](https://jdunham.io/dgo/reference/dgo_fit-methods.html). By default,
this method returns post-warmup samples for the `theta_bar` parameters, but like
other methods takes a `pars` argument.

```{r}
head(as.data.frame(dgmrp_out_abortion))
```

To poststratify the results use `poststratify()`. Here, we use the group
population proportions bundled as `annual_state_race_targets` to reweight and
aggregate estimates to strata defined by state-years.

```{r}
poststratify(dgmrp_out_abortion, annual_state_race_targets, strata_names =
  c("state", "year"), aggregated_names = "race3")
```

To plot the results use `dgirt_plot()`. This method plots summaries of posterior
samples by time period. By default, it shows a 95% credible interval around
posterior medians for the `theta_bar` parameters, for each local geographic
area. Here we omit the CIs.

```{r dgmrp_plot}
dgirt_plot(dgmrp_out_abortion, y_min = NULL, y_max = NULL)
```

`dgirt_plot()` can also plot the `data.frame` output from `poststratify()`,
given arguments that identify the relevant variables. Below, we aggregate over
the demographic grouping variable `race3`, resulting in a `data.frame` of
estimates by state-year. 

```{r dgmrp_plot_ps}
ps <- poststratify(dgmrp_out_abortion, annual_state_race_targets, strata_names =
  c("state", "year"), aggregated_names = "race3")
head(ps)
dgirt_plot(ps, group_names = NULL, time_name = "year", geo_name = "state")
```

In the call to `dgirt_plot()`, we passed the names of the `state` and `year`
variables. The `group_names` argument was then `NULL`, because there were no
grouping variables left after we aggregated over `race3`.

