---
title: "Policy Liberalism"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates estimation of latent policy liberalism from
individuals' responses to five survey items, using the Bayesian group-level IRT
model implemented in `dgirt()`.

```{r, knitr-options, echo = FALSE, result = 'hide', message = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  cache = TRUE,
  comment = "#>")
library(dgo)
```

## Prepare input data

`shape()` prepares input data for use with the modeling functions `dgirt()` and
`dgmrp()`. Here we use the included `opinion` dataset.

```{r}
dgirt_in_liberalism <- shape(opinion, item_names = c("abortion",
    "affirmative_action","stemcell_research" , "gaymarriage_amendment",
    "partialbirth_abortion") , time_name = "year", geo_name = "state",
  group_names = "race3", geo_filter = c("CA", "GA", "LA", "MA"))
```

In this call to `shape()` we specified:

* the survey item response variables as `item_names`
* which variable represents time (`year`), since dgo models are dynamic;
* the variables representing respondent characteristics (`state` and `race3`),
  because dgo models are group-level.

Notice that we named only one of these variables defining respondent groups
using the `group_names` argument. The `geo_name` argument always takes the
variable giving respondents' local geographic area; it will be modeled
differently.

Using the argument `geo_filter`, we subset the input data to the given values of
the `geo_name` variable. And with the `id_vars` argument, we named an identfier
that we'd like to keep in the processed data. (Other unused variables will be
dropped.)

## Inspect the result

`summary()` gives a high-level description of the result.

```{r}
summary(dgirt_in_liberalism)
```

`get_n()` and `get_item_n()` give response counts.

```{r}
get_n(dgirt_in_liberalism, by = "state")
get_item_n(dgirt_in_liberalism, by = "year")
```

## Fit a model

`dgirt()` estimates a latent variable based on responses to multiple survey
questions.  Here, we'll use it to estimate latent policy liberalism over time,
for the groups defined by `state` and `race3`. (Specifically, by their Cartesian
product.)

Under the hood, `dgirt()` uses RStan for MCMC sampling, and arguments can be
passed to RStan's
[`stan()`](https://www.rdocumentation.org/packages/rstan/versions/2.15.1/topics/stan)
via the `...` argument of `dgirt()`. This is almost always desirable. Here, we
specify the number of sampler iterations, chains, and cores.

```{r, warning = FALSE, message = FALSE, results = 'hide'}
dgirt_out_liberalism <- dgirt(dgirt_in_liberalism, iter = 3000, chains = 4,
  cores = 4, seed = 42)
```

The model results are held in a `dgirt_fit` object. Methods from RStan like
`extract()` are available if needed because `dgirt_fit` is a subclass of
`stanfit`.  But dgo provides its own methods for typical post-estimation tasks.

## Work with results

For a high-level summary of the result, use `summary()`.

```{r}
summary(dgirt_out_liberalism)
```

To apply scalar functions to posterior samples, use
[`summarize()`](https://jdunham.io/dgo/reference/dgo_fit-methods.html). The
default output gives summary statistics for the model's `theta_bar` parameters,
which represent group means. These are indexed by time (`year`) and group, where
groups are again defined by local geographic area (`state`) and any other
respondent characteristics (`race3`).

```{r}
head(summarize(dgirt_out_liberalism))
```

Alternatively,
[`summarize()`](https://jdunham.io/dgo/reference/dgo_fit-methods.html) can apply
arbitrary functions to posterior samples for whatever parameter is given by its
`pars` argument.

```{r}
summarize(dgirt_out_liberalism, pars = "xi", funs = "var")
```

To access posterior samples in tabular form use
[`as.data.frame()`](https://jdunham.io/dgo/reference/dgo_fit-methods.html). By
default, this method returns post-warmup samples for the `theta_bar` parameters,
but like other methods takes a `pars` argument.

```{r}
head(as.data.frame(dgirt_out_liberalism))
```
To poststratify the results use `poststratify()`. Here, we use the group
population proportions bundled as `annual_state_race_targets` to reweight and
aggregate estimates to strata defined by state-years.

```{r}
poststratify(dgirt_out_liberalism, annual_state_race_targets, strata_names =
  c("state", "year"), aggregated_names = "race3")
```

To plot the results use `dgirt_plot()`. This method plots summaries of posterior
samples by time period. By default, it shows a 95% credible interval around
posterior medians for the `theta_bar` parameters, for each local geographic
area. Here we omit the CIs.

```{r dgirt_plot}
dgirt_plot(dgirt_out_liberalism, y_min = NULL, y_max = NULL)
```

`dgirt_plot()` can also plot the `data.frame` output from `poststratify()`,
given arguments that identify the relevant variables. Below, we aggregate over
the demographic grouping variable `race3`, resulting in a `data.frame` of
estimates by state-year. 

```{r dgirt_plot_ps}
ps <- poststratify(dgirt_out_liberalism, annual_state_race_targets, strata_names
  = c("state", "year"), aggregated_names = "race3")
head(ps)
dgirt_plot(ps, group_names = NULL, time_name = "year", geo_name = "state")
```

In the call to `dgirt_plot()`, we passed the names of the `state` and `year`
variables. The `group_names` argument was then `NULL`, because there were no
grouping variables left after we aggregated over `race3`.
